---
title: 'Data Science Project'
output: html_document
author: "Jason Eubank"
date: "`r format(Sys.Date(), '%m/%d/%y')`"
---

```{r set options, include=FALSE}
# DO NOT CHANGE THE LINE BELOW 
knitr::opts_chunk$set(echo = TRUE)
```

``` {css styling, echo=FALSE}

<style>
.tocify {
max-width: 175px !important;
}
</style>

<style>
.main-container {
width: 100%;
max-width: 940px;
margin-left: 250px;
margin-right: auto;
}
</style>

<style>
.red-header {
  color: red;
}
</style>

```

```{r logo, echo = FALSE}

htmltools::img(src = 'https://upload.wikimedia.org/wikipedia/en/0/03/National_Basketball_Association_logo.svg',
                height = '250px',
                alt = 'logo',
                style = 'position: fixed; top: -40px; left: 5px;')
```


# Introduction  

The purpose of this project is to analyze how professional basketball team schedules influence performance. Using data from recent seasons, it explores how schedule congestion, travel demands, and home/away balance affect teams over time. All analysis was completed using tidyverse, ggplot2, and plotly. The provided data was used as truth (for example, some "home" games have been played at secondary locations, including TOR's entire 2020-21 season. These are not reflected in the data, so it is not accounted for) Note that the 2024-25 schedules in `schedule_24_partial.csv` intentionally include only 80 games, as the league holds 2 games out for each team in the middle of December due to unknown NBA Cup matchups.    

**Note:**    
**Throughout this document, any `season` column represents the year each season started. For example, the 2015-16 season will be in the dataset as 2015. We may refer to a season by just this number (e.g. 2015) instead of the full text (e.g. 2015-16).**   


# Setup and Data    

```{r load data, message = F, warning = F}
library(tidyverse)
schedule <- read_csv("~/Downloads/schedule.csv")
draft_schedule <- read_csv("~/Downloads/schedule_24_partial.csv")
locations <- read_csv("~/Downloads/locations.csv")
game_data <- read_csv("~/Downloads/team_game_data.csv")
```

## Part 1 -- Schedule Congestion Analysis               


#The first step identifies periods where your selected team plays four games in six nights. (Note: The stretches can overlap)     
#OKC is team chosen, any team can be inserted. Apply to 2024-25 schedule.

```{r}
draft_schedule_OKC <- draft_schedule %>%
  filter(team == "OKC")

count_4in6<- function(data, date_col){
  data %>%
    # Standarize the data 
    transmute(gamedate = as.Date({{date_col}})) %>% 
    arrange(gamedate) %>%
    mutate(n_games = purrr::map_int(gamedate, ~ sum(gamedate >= .x & gamedate <= .x+5))) %>%
    filter(n_games >= 4) %>%
  mutate(window_end = gamedate + 5) %>%
  select(window_start = gamedate, window_end, n_games)
}
res <- count_4in6(draft_schedule_OKC, gamedate)
res
nrow(res)

```
## Part 2    

#Apply to find historical four in six averages 
  

```{r}
by_team_season <- schedule %>%
  filter(season >= 2014, season <= 2023) %>%
  group_by(team,season) %>%
  summarise(games_played = n(),
            four_in_six = nrow(count_4in6(pick(gamedate), gamedate)), 
            .groups = "drop"
  ) %>%
  mutate(four_in_six_per82 = four_in_six * 82 / games_played)
overall_avg_q2 <- by_team_season %>% 
  summarise(avg_4in6_per82 = mean(four_in_six_per82, na.rm = TRUE))
overall_avg_q2

```

#Locate highest average number of 4-in-6 stretches between 2014-15 and 2023-24. Identify which team has had the highest and lowest average.


```{r}
by_team_season <- schedule %>%
  filter(season >= 2014, season <= 2023) %>%
  group_by(team,season) %>%
  summarise(games_played = n(),
            four_in_six = nrow(count_4in6(pick(gamedate), gamedate)), 
            .groups = "drop") %>%
  mutate(four_in_six_per82 = four_in_six * 82 / games_played)

#Avg per team across seasons 
team_avgs <- by_team_season %>%
  group_by(team) %>%
  summarise(avg_4in6_per82 = mean(four_in_six_per82, na.rm = TRUE)) %>%
  arrange(desc(avg_4in6_per82))

#Identify highest and lowest 
highest <- team_avgs %>% slice(1)
lowest  <- team_avgs %>% slice(n()) 

highest 
lowest



``` 
  

The difference between the most CHA (28.1) and least NYK(22.2) is nearly 6 stretches per 82 games. This difference is likely the result of chance since it is fairly small compared to the amount of teams in the league and the 10 seasons considered.
  


## Part 3 -- Trends and Visualizations                   


#Locate Trends


```{r}
#B2B Trends 
b2b_trend <- schedule %>%
mutate(gamedate = as.Date(gamedate)) %>%
  arrange(team, gamedate) %>%
  group_by(team, season) %>%
  mutate(prev_game = lag(gamedate), b2b = (gamedate - prev_game == 1)) %>%
  summarise(b2b_total = sum(b2b, na.rm = TRUE), .groups = "drop") %>% group_by(season) %>%
  summarise(avg_b2b_per_team = mean(b2b_total), .groups = "drop")

#Visualization 
library(ggplot2)

ggplot(b2b_trend, aes( x = season, y = avg_b2b_per_team)) + 
  geom_line() +
  geom_point() +
  labs(
    title = "Average Back to Backs per Team by Season",
    x = "Season",
    y = "Avg B2B per Team"
  ) + 
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

``` 

```{r}

road_trips_team_season <- schedule %>% 
  mutate(gamedate = as.Date(gamedate)) %>%
  arrange(team, gamedate) %>%
  group_by(team, season) %>%
  #Flag Away Games and  ID each trip 
  mutate(
    away = (home == 0) ,
    prev_away = c(FALSE, away[-length(away)]) ,
    new_trip = away & (!prev_away) , 
    trip_id = cumsum(new_trip)
  ) %>%
  
#Filter Away Games 
  filter(away) %>% 
  group_by(team, season, trip_id) %>% 
  
#Calculate Length and Count 3+ Stretches
  summarise(trip_len = n(), .groups = "drop") %>%
  group_by(team, season) %>%
  summarise(n_3plus_road_trips = sum(trip_len >= 3), .groups = "drop")

#Visualization
road_trip_trend <- road_trips_team_season %>%
  group_by(season) %>%
  summarise(avg_3plus_road_trips = mean(n_3plus_road_trips), .groups = "drop")
ggplot(road_trip_trend, aes(x = season, y = avg_3plus_road_trips)) + 
  geom_line() +
  geom_point() +
  labs(title = "Average 3+ Road Game Stretches by Team per Season", 
       x     = "Season",
       y     = "Avg 3+ Road Game Stretches per Team") +
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_y_continuous(limits = c(0,10)) #Widen scale

```
 

From 2014-15 to 2023-24, the average number of back to back games per team has steadily declined. On average teams were player around 18 to 19 back to back per season in early years, but that number has dropped to around 14 in recent years. It seems the league reduced the amount of back to backs played each year. This can help reduce schedule congestion and prioritize player health.

Looking at road stretches with 3 or more games, the trend has been relatively stable, averaging between 5 and 7 per team per season. While there is some year to year variation, the long term pattern is consistent. The league appears to balance these trips for travel efficiency. 

### Question 7    

#Visualizations

```{r}
library(plotly)

#Filter 
okc_den <- draft_schedule %>%
  filter(team == "OKC" | team == "DEN") %>%
  mutate(gamedate = as.Date(gamedate)) %>%
  arrange(team, gamedate) %>%
  group_by(team) %>%
  
#Track Previous Game Date
mutate(
  prev_date = c(NA, gamedate[-length(gamedate)]),
#Track Next Game Date  
  next_date = c(gamedate[-1], NA),
#B2B
  b2b_any   = (!is.na(prev_date) & (gamedate-prev_date == 1))|
              (!is.na(next_date) & (next_date - gamedate == 1)),
#Home or Away
home_fac = ifelse(home == 1, "Home", "Away"),

#Hover for Plotly  
hover_txt = paste0(
    "Team: ", team,
    "<br>Date:", gamedate,
    "<br>Opponent: ", opponent,
    "<br>Venue: ", home_fac,
    "<br>B2B: ", ifelse(b2b_any, "Yes", "No")
  )
) %>%
ungroup()

p <- ggplot(okc_den, aes(x = gamedate,
                         y = 1,
                         color = home_fac,
                         shape = b2b_any,
                         text = hover_txt)) +
  geom_point( size = 3) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 17), name = "Back to Back") +
  scale_y_continuous(NULL, breaks = NULL) +
  labs(
    title = "2024-25 OKC/DEN Schedule",
    x = "Date",
    y = NULL,
    color = "Venue",
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()) +
  facet_wrap(~ team, ncol = 1, scales = "free_y")

gp <- ggplotly(p, tooltip = "text") %>%
  layout(
    legend = list(orientation = "h", y = -0.15),
    xaxis = list(rangeslider = list(visible = TRUE))
  )

gp

```
 

#Locate the best and worst part of OKC’s 2024-25 draft schedule.


The best part of OKC's 2024-25 draft schedule is from 03/27/25 to 04/02/25, when the team plays 4 straight home games with no back to back games during the stretch. This gives the opportunity to get valuable rest and gear up for a playoff run. 

The toughest stretch occurs from 02/21 to 03/03, where OKC plays 7 games in 11 days, including two different back to backs Feb(23-24) and Mar(2-3). This period has disruptive travel with 5 away games and the only 2 home games coming on the back end of back to backs. This creates a significant challenge to catch any rhythm or gain any rest. 

## Part 3 -- Modeling     
  

#Determine Most Hurt and Most Helped by schedule 

    

```{r}
sched_feat <- schedule %>%
  mutate(gamedate = as.Date(gamedate),
         away = (home == 0)) %>%
  filter(season >= 2019, season <= 2023) %>%
  arrange(team, gamedate) %>%
  group_by(team, season) %>%
  mutate(
    prev_date = lag(gamedate),
    next_date = lead(gamedate),
    
#B2B
    b2b_any = (!is.na(prev_date) & (gamedate - prev_date == 1)) |
              (!is.na(next_date) & (next_date - gamedate == 1)),

#Count 4in6
n_in_6 = purrr::map_int(gamedate, \(d) sum(gamedate >= d & gamedate <= d + 5)),
four_in_six = (n_in_6 >= 4),

#Road Trips 
prev_away = lag(away, default = FALSE),
new_trip  = away & !prev_away,
trip_id   = ifelse(away,cumsum(new_trip), NA_integer_)
) %>%
group_by(team, season, trip_id) %>%
mutate(road_trip_index = ifelse(away, row_number(), 0L)) %>%
ungroup() %>%
mutate(
  b2b             = as.integer(b2b_any),
  four_in_six     = as.integer(four_in_six),
  road_trip_index = replace_na(road_trip_index, 0L),
  home            = as.integer(home),
  win             = as.integer(win)
) %>%
select(season, team, gamedate, home, away, win, b2b, four_in_six, road_trip_index)

#Fit model 

sched_scored <- sched_feat %>%
  filter(season >= 2019, season <= 2023)

sched_model <- glm(
  win ~ b2b + four_in_six + road_trip_index + home,
  data = sched_scored,
    family = binomial()
) 
summary(sched_model)

sched_scored$pred_win_prob <- predict(sched_model, newdata = sched_scored, type = "response")

sched_scored <- sched_scored %>%
  mutate(expected_wins = ifelse(pred_win_prob > 0.50, 1, 0))

#Actual vs Expected 
  team_summary <- sched_scored %>%
    group_by(team) %>%
    summarise(
      total_actual_wins   = sum(win, na.rm = TRUE),
      total_expected_wins = sum(expected_wins, na.rm = TRUE),
      diff          = total_actual_wins - total_expected_wins,
      .groups       = "drop"
    ) %>%
  arrange(diff)

most_hurt   <- slice_head(team_summary, n = 1)
most_helped <- slice_tail(team_summary, n = 1)

most_helped
most_hurt
```

- Most Helped by Schedule: MIL (260 wins)     
- Most Hurt by Schedule: DET (94 wins)     

I fit a logistic regression model on games from 2019 to 2023 with wins as the outcome and schedule features as predictors. The estimates indicate that back to backs are influential in reducing win probability, while home court advantage increases a team's chances to win. The four in six and road trip indicators are small so they are not significant to the outcome. Overall, the model points to venue and back to back density as the schedule factors with clear impact. 







